---
title: "Phosphorus Budgets for FORB"
author: "Kenneth Anderson"
date: "2023-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("dplyr")
```

## ODNR Budgets: 

ODNR makes budgets for P reduction in potential wetlands using the following method: 

1. Put the lat/long of the outlet of the wetland into USGS Stream Stats to get the drainage area of the wetland (convert to acres)

2. Based on the size of the wetland correct for the amount of the drainage the wetland interacts with
  a. 0-99 acres: 95%
  b. 100-999 acres: 75%
  c. 1,000-9,999 acres: 25%
  d. 10,000-99,999 acres: 10%
  e. 100,000-999,999 acres: 1%
  f. 1,000,000 acres +: 0.5%

```{r}
# Drainage Area from Stream Stats
DA = 63
# Treated drainage area (corrected for drainage size)
tDA = if(DA <= 99){DA*0.95} else if(DA >= 100 & DA <= 999){DA*0.75} else if(DA >= 1000 & DA <= 9999){DA*0.25} else if(DA >= 10000 & DA <= 99999){DA*0.1} else if(DA >= 100000 & DA <= 999999){DA*0.01} else if(DA > 100000000){DA*0.005}
# units = acres
print(tDA)
```

3. Estimate P loading rate using the USGS SPARROW model (units are pounds P / year) based on total drainage area treated (calculated in step 2)

```{r}
# P Loading Rates for Ohio from USGS SPARROW:
# HUC-8 Drainage basin ID
HUC8 = c(4100001,4100008,4100009,4100010,4100011,4100003,4100005,5120101,5040006,5090202,4120200,4120101,4100007,4100006,5040001,5080001,5060001,4100014,4110001,4110002,4110003,4110004)
# Total phosphorus yield for each basin (kg/km^2/year)
TPyieldkgkm=c(75.2,131,129,83.9,140,114,106,293,206,216,0,123,152,114,134,201,209,149,179,249,245,109)
loading=data.frame(HUC8,TPyieldkgkm)
# correct to lbs/acre/year from kg/km^2/year:
corr=112.0850759
loading$TPyield = loading$TPyieldkgkm/corr

# select the HUC-8 basin that the wetland is a part of 
loadFORBHUC = loading$TPyield[loading$HUC8 == 4100005]
# multiply by the treated drainage area from step 2
loadFORB = loadFORBHUC * tDA
# units = lbs / year
print(loadFORB)
```

4. Estimate P removal based on literature (as proposed by TetraTech; units: pounds P/ acre / year): 
  a. Low estimate: 18% removal of P loading
  b. Hybrid estimate: 64% removal of P loading
  c. High estimate: 78% of P loading

```{r}
# P Removal Rates:
low = loadFORB*0.18
hybrid = loadFORB*0.64
high = loadFORB*0.78

# units = lbs/year
print(low)
print(hybrid)
print(high)
```

  
4b. For floodplain wetlands estimates are based on constants from Noe et al. 2019: 3.2 pounds P/ acre / year removal

### ***ASSUMPTIONS IN ODNR BUDGETS:*** 
  Not that many, but boy are they big assumptions<br>
  1. Drainage area is accurate (it's not)<br>
  2. SPARROW gives an accurate measurement of P loading<br>
  3. Literature estimates of loading are accurate <br>
  * Honestly this probably isn't the worst assumption if only because their estimates are so wide ranging.<br>
  4. Seasonal variation does not matter to whoever is looking for this budget. <br>
  5. Annual variation does not play a major role in phosphorus loading, or retention, or hydrologic dynamics.<br>

## Hybrid Modelling Budget 

Calculate phosphorus loading as above:

```{r}
# units = lbs / year
print(loadFORB)
```

Now we'll calculate the P retention at a couple locations across the wetland. <br>

We'll start with Wetland Complex 4:
```{r}
# Add in the P dataset (units here are mg/L for TP)
setwd("C:/Users/kande120/OneDrive - Kent State University/Phosphorus Budgets/FORB/Data")
Pdata<-read.csv("FORB P Budget Full inletsandoutlets.csv")
# subset out event sampling 
Pdata<- subset(Pdata, Sampling == "baseflow")
# Calculate percent retention in Wetland Complex 4
Pdata$retainedPcomp4=Pdata$Inlet_wetland_FORB_9-Pdata$Outlet_wetland_FORB_12
Pdata$percentPcomp4 = Pdata$retainedPcomp4 / Pdata$Inlet_wetland_FORB_9 * 100
percentRemoval = mean (Pdata$percentPcomp4,na.rm=T)
print(percentRemoval)
```
We get an average percent removal of 8.7% but see here how variable that is: 

```{r}
# make it realize it's a date time value
Pdata$Date <- as.Date(Pdata$Date, format="%m/%d/%Y")
# make the plot
library(ggplot2)
p1<-ggplot(data=Pdata, aes(x=Date, y=percentPcomp4)) +
  geom_bar(stat="identity")+scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y")+theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(p1)
```

Now let's try out the whole flowpath, starting at the tile draining into Wetland Complex 4 and ending in the two streams that deposit into the Maumee:
```{r}
# take the average outflow between streams 1 and 2 (for samples at the end of the stream)
Pdata$outflowTP <- rowMeans(Pdata[c('Outlet_stream1_FORB_17_END', 'Outlet_stream2_FORB_18')], na.rm=TRUE)
# inflow remains largely the same for the main flowpath because the inlet tile is not sampled often
Pdata$inflowTP <- rowMeans(Pdata[c('Inlet_wetland_FORB_9', 'Inlet_tile_FORB_15')], na.rm=TRUE)

# Calculate percent retention in whole wetland
Pdata$retainedP=Pdata$inflowTP-Pdata$outflowTP
Pdata$percentP = Pdata$retainedP / Pdata$inflowTP * 100
percentRemoval = mean (Pdata$percentP,na.rm=T)
print(percentRemoval)
```
We want to still look at the variability of this measurement: 
```{r}
# make it realize it's a date time value
Pdata$Date <- as.Date(Pdata$Date, format="%m/%d/%Y")
# make the plot
library(ggplot2)
p2<-ggplot(data=Pdata, aes(x=Date, y=percentP)) +
  geom_bar(stat="identity")+scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y")+theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(p2)
```

We see that the full wetland much more consistently retains P than just wetland complex 4. <br>

P retention at the wetland ranges from 19 to 83 % retention, but that doesn't account for seasonal differences so lets try to take annual averages: 

```{r}
# 2021
Pdata2021<- subset(Pdata, Year == "2021")
percentRemoval2021 = mean (Pdata2021$percentP,na.rm=T)
print(percentRemoval2021)
# 2022
Pdata2022<- subset(Pdata, Year == "2022")
percentRemoval2022 = mean (Pdata2022$percentP,na.rm=T)
print(percentRemoval2022)
# 2023
Pdata2023<- subset(Pdata, Year == "2023")
percentRemoval2023 = mean (Pdata2023$percentP,na.rm=T)
print(percentRemoval2023)
```
This gives us a range from 42% retention to 62% retention (but our data is not evenly distributed across different years) <br>

We can now apply these measurements of P retention to the estimated phosphorus loading from SPARROW modelling. 

```{r}
# units = lbs / year
print(loadFORB)

# P Removal Rates:
low = loadFORB*0.42
high = loadFORB*0.62

# units = lbs/year
print(low)
print(high)
```
By this budgetting we can estimate that Forder Bridge retains between 24 and 35 lbs P / year

### Rainfall at Ft. Wayne Budget
We'll use data from Ft. Wayne Indiana (31 miles away) downloaded from Weather Underground.<br>

We'll calculating runoff using the curve number of the soils: 
We will calculate the runoff that flows through the wetland during rain events based on some information about soil properties.


$$Q=\begin{cases}
0 \text{ for } P \le I_a\\
{(P-I_a)^2 \over P-I_a+S} \text{ for } P > I_a
\end{cases}$$

where: <br>
Q = runoff (inches)<br>
P = rainfall (inches)<br>
S = potential maximum soil moisture retention after runoff begins (inches)<br>
I_a = initial abstraction (inches; this is the amount of rainfall needed for runoff to occur)<br>



The runoff curve (CN) is used to calculate S
$$S = {1000 \over CN} -10$$
We will assume that: $$ I_a = 0.2S$$ as a standard correction. 
(See here https://onlinelibrary.wiley.com/doi/10.1111/j.1752-1688.2006.tb04481.x for discussion of whether that is a good correction but we're going to power through with it)


CN will depend on the soil of the site. In this case it's probably around 80, fairly poorly drained, but we may want to come back and think abit more about what our curve number should be. <br>

```{r}
# Drainage Area calculated by StreamStats
tDA = 59.9
# convert tDA to m^2
tDAm2=tDA*4046.86
# units = m^2
print(tDAm2)
```

Then we import data on precipitation from weather underground at the Ft. Wayne weather station (31 miles away)

```{r}
setwd("C:/Users/kande120/OneDrive - Kent State University/Phosphorus Budgets/FORB/Data")
raindata<-read.csv("Ftwaynerainfall.csv")
# make it recognize that it's a date-time value
raindata = mutate(raindata, datetime = as.POSIXct(datetime, format = "%m/%d/%Y"))
# Rename so our old code keeps working 
dayrain=raindata
```
Then we calculate runoff as above.
```{r}
# Start out by calculating S (inches)
CN=85
S = (1000/CN)-10
# Calculate I (inches)
Ia = 0.2*S
# don't need to convert rain to inches/day, but lets name it so our old code still works
dayrain$precip = dayrain$precipaccum_in
# caculate runoff (inches) for each day in the dataset
dayrain$Q = ifelse(dayrain$precip<=Ia, 0, ((dayrain$precip-Ia)^2)/(dayrain$precip-Ia+S))
# scale up to the entire drainage area
dayrain$Q=tDAm2*dayrain$Q
# convert back to meters 
dayrain$Qm = dayrain$Q * 0.0254
# convert discharge to liters
dayrain$QL = dayrain$Qm * 1000
# Sum up data to get monthly values
library(lubridate)
library(tidyverse)
dayrain$Date=as.POSIXct(dayrain$datetime,format="%Y-%m-%d")
# Summarize the dataset by month and year
monthrain=dayrain %>% mutate(Month=month(Date,label = T),Year=year(Date)) %>%
  group_by(Year,Month) %>% summarise(QL=sum(QL,na.rm=T))
```

Now we add in the monthly measurements of Phosphorus concentrations to calculate P loading.
```{r}
# Add in the P dataset (units here are mg/L for TP)
setwd("C:/Users/kande120/OneDrive - Kent State University/Phosphorus Budgets/FORB/Data")
Pdata<-read.csv("FORB P Budget Full inletsandoutlets.csv")
# subset out event sampling 
Pdata<- subset(Pdata, Sampling == "baseflow")
# take the average outflow between streams 1 and 2 (for samples at the end of the stream)
Pdata$outflowTP <- rowMeans(Pdata[c('Outlet_stream1_FORB_17_END', 'Outlet_stream2_FORB_18')], na.rm=TRUE)
# Average the inflow between the main inlet tile and the start of complex 4 (the main inlet tile is not sampled as often as the start of complex 4)
Pdata$inflowTP <- rowMeans(Pdata[c('Inlet_wetland_FORB_9', 'Inlet_tile_FORB_15')], na.rm=TRUE)
# merge the datasets
total <- merge(Pdata,monthrain,by=c("Month","Year"),all=TRUE)
# Measure the mass of P entering the wetland (units = mg P/month)
total$Pin=total$inflowTP*total$QL
# calculate the total mass of P exiting the wetland (units = mg P/month)
total$Pout=total$outflowTP*total$QL
# Convert to lbs P/month
total$Pinlb=total$Pin/453592
total$Poutlb=total$Pout/453592
# calculate the amount of P removed
total$Premoved=total$Pinlb-total$Poutlb
```

Now we need to look at where we have data and attempt to extrapolate where we do not
```{r}
# To start lets look at just 2023
total23<- subset(total, Year == "2023")
total23<- subset(total23, select=c(Month,Year,Pinlb,Poutlb,Premoved,QL,inflowTP,outflowTP))
print(total23)
```

We can assume there is no P retained when there is no discharge: 
```{r}
library(dplyr)
# first make all NaNs into NAs
total23 <- total23 %>% mutate_all(~ifelse(is.nan(.), NA, .))
# now replace NAs where discharge = 0 as also 0
total23$Premoved=ifelse(test = total23$QL == 0, yes = "0", no = total23$Premoved)
```
We have one event with loading data but no retention data. Here we can estimate retention based on the average retention of the other months. 
```{r}
# tell r that Premoved is numeric: 
total23$Premoved=as.numeric(total23$Premoved)
# Calculate percent removal
total23$percentP=total23$Premoved/total23$Pinlb
percentRemoval = mean (total23$percentP,na.rm=T)
print(percentRemoval)
```
There's a pretty good amount of variation here, but lets power through and use the average. 
```{r}
# replace the month in question (July 2023)
total23$Premoved=ifelse(test = total23$Month == "Jul" & total23$Year == "2023", yes = total23$Pinlb*percentRemoval, no = total23$Premoved)
```

Now we can sum up the total P removed by the wetland over the course of the year: 
```{r}
Removal = sum (total23$Premoved,na.rm=T)
# This is the removal of P by pool by over the course of the 8 months that we have data for
print(Removal)
```
We can estimate the other 4 months by estimating a low and a high value for P loading and P retention. <br> 
We have discharge data for all months, so we don't want to just estimate loading, when we can estimate concentration instead:
```{r}
# our low estimate is the lowest concentration we saw: 
lowPcon = min(total23$inflowTP,na.rm = TRUE)
# our high estimate is the highest concentration we saw: 
highPcon= max(total23$inflowTP,na.rm = TRUE)

# then we can estimate the low end of % retention
lowPper = min(total23$percentP,na.rm = TRUE)
# and then the high end of % retention
highPper = max(total23$percentP,na.rm = TRUE)


# now calculate P load and retention using these concentrations to estimate the high and low end of P removal by the wetland:

# first for the low estimate
total23$Premovedlow=ifelse(test = is.na(total23$Premoved) , yes = total23$QL*lowPcon/453592*lowPper, no = total23$Premoved)
# then for the high estimate
total23$Premovedhigh=ifelse(test = is.na(total23$Premoved) , yes = total23$QL*highPcon/453592*highPper, no = total23$Premoved)

# Now we can sum up our full months: 
lowRemoval = sum (total23$Premovedlow,na.rm=T)
highRemoval = sum (total23$Premovedhigh,na.rm=T)
print(lowRemoval)
print(highRemoval)
```

 
### ***ASSUMPTIONS IN THIS BUDGET:*** 
  1. Drainage area is accurate (it's not)<br>
  2. Curve number is a good choice for this site <br>
  3. Groundwater dynamics are captured <br>
  4. Tile drains are accounted for (they're not) <br>
  5. Nearby rainfall data is accurate for the site <br>

