---
title: "Phosphorus Budgets for SJRE"
author: "Kenneth Anderson"
date: "2023-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("dplyr")
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## ODNR Budgets: 

ODNR makes budgets for P reduction in potential wetlands using the following method: 

1. Put the lat/long of the outlet of the wetland into USGS Stream Stats to get the drainage area of the wetland (convert to acres)

2. Based on the size of the wetland correct for the amount of the drainage the wetland interacts with
  a. 0-99 acres: 95%
  b. 100-999 acres: 75%
  c. 1,000-9,999 acres: 25%
  d. 10,000-99,999 acres: 10%
  e. 100,000-999,999 acres: 1%
  f. 1,000,000 acres +: 0.5%

```{r}
# Drainage Area from Stream Stats
DA = 48
# Treated drainage area (corrected for drainage size)
tDA = if(DA <= 99){DA*0.95} else if(DA >= 100 & DA <= 999){DA*0.75} else if(DA >= 1000 & DA <= 9999){DA*0.25} else if(DA >= 10000 & DA <= 99999){DA*0.1} else if(DA >= 100000 & DA <= 999999){DA*0.01} else if(DA > 100000000){DA*0.005}
# units = acres
print(tDA)
```

3. Estimate P loading rate using the USGS SPARROW model (units are pounds P / year) based on total drainage area treated (calculated in step 2)

```{r}
# P Loading Rates for Ohio from USGS SPARROW:
# HUC-8 Drainage basin ID
HUC8 = c(4100001,4100008,4100009,4100010,4100011,4100003,4100005,5120101,5040006,5090202,4120200,4120101,4100007,4100006,5040001,5080001,5060001,4100014,4110001,4110002,4110003,4110004)
# Total phosphorus yield for each basin (kg/km^2/year)
TPyieldkgkm=c(75.2,131,129,83.9,140,114,106,293,206,216,0,123,152,114,134,201,209,149,179,249,245,109)
loading=data.frame(HUC8,TPyieldkgkm)
# correct to lbs/acre/year from kg/km^2/year:
corr=112.0850759
loading$TPyield = loading$TPyieldkgkm/corr

# select the HUC-8 basin that the wetland is a part of 
loadSJREHUC = loading$TPyield[loading$HUC8 == 4100003]
# multiply by the treated drainage area from step 2
loadSJRE = loadSJREHUC * tDA
# units = lbs / year
print(loadSJRE)
```

4. Estimate P removal based on literature (as proposed by TetraTech; units: pounds P/ acre / year): 
  a. Low estimate: 18% removal of P loading
  b. Hybrid estimate: 64% removal of P loading
  c. High estimate: 78% of P loading

```{r}
# P Removal Rates:
low = loadSJRE*0.18
hybrid = loadSJRE*0.64
high = loadSJRE*0.78

# units = lbs/year
print(low)
print(hybrid)
print(high)
```

  
4b. For floodplain wetlands estimates are based on constants from Noe et al. 2019: 3.2 pounds P/ acre / year removal

### ***ASSUMPTIONS IN ODNR BUDGETS:*** 
  Not that many, but big assumptions<br>
  1. Drainage area is accurate (it's not)<br>
  2. SPARROW gives an accurate measurement of P loading<br>
  3. Literature estimates of loading are accurate <br>
  * Honestly this probably isn't the worst assumption if only because their estimates are so wide ranging.<br>
  4. Seasonal variation does not matter to whoever is looking for this budget. <br>
  5. Annual variation does not play a major role in phosphorus loading, or retention, or hydrologic dynamics.<br>

## Calculating Runoff from Rainfall
Calculate drainage area and treated drainage area using digital elevation models:

```{r}
# Drainage Area calculated by Bishwodeep using Digital Elevation Model (DEM)
tDA = 103.32
# convert tDA to m^2
tDAm2=tDA*4046.86
# units = m^2
print(tDAm2)
```
Then we import data on precipitation from the sensor network

```{r}
setwd("C:/Users/kande120/OneDrive - Kent State University/Phosphorus Budgets/SJRE/Data")
raindata<-read.csv("SJREsensors.csv")
# These are the first 6 rows to give you an idea what the datatable looks like
head(raindata)
# make it recognize that it's a date-time value
raindata = mutate(raindata, datetime = as.POSIXct(datetime, format = "%Y-%m-%d %H:%M:%S"))

# Here we will aggregate our rainfall values so we have a sum of rainfall for each day
DF2 <- transform(raindata, Date = as.Date(datetime))
dayrain=aggregate(precipaccum ~ Date, DF2, sum)
```

We will calculate the runoff that flows through the wetland during rain events based on some information about soil properties.


$$Q=\begin{cases}
0 \text{ for } P \le I_a\\
{(P-I_a)^2 \over P-I_a+S} \text{ for } P > I_a
\end{cases}$$

where: <br>
Q = runoff (inches)<br>
P = rainfall (inches)<br>
S = potential maximum soil moisture retention after runoff begins (inches)<br>
I_a = initial abstraction (inches; this is the amount of rainfall needed for runoff to occur)<br>





The runoff curve (CN) is used to calculate S
$$S = {1000 \over CN} -10$$
We will assume that: $$ I_a = 0.2S$$ as a standard correction. 
(See here https://onlinelibrary.wiley.com/doi/10.1111/j.1752-1688.2006.tb04481.x for discussion of whether that is a good correction but we're going to power through with it)


CN will depend on the soil of the site. In this case it's probably around 80, fairly poorly drained (we know that the surrounding fields flood often). <br>
```{r}
# Start out by calculating S (inches)
CN=80
S = (1000/CN)-10
# Calculate I (inches)
Ia = 0.2*S
# convert rain to inches/day
dayrain$precip = dayrain$precipaccum/25.4
# caculate runoff (inches) for each day in the dataset
dayrain$Q = ifelse(dayrain$precip<=Ia, 0, ((dayrain$precip-Ia)^2)/(dayrain$precip-Ia+S))
# scale up to the entire drainage area
dayrain$Q=tDAm2*dayrain$Q
# convert back to meters 
dayrain$Qm = dayrain$Q * 0.0254
# convert discharge to liters
dayrain$QL = dayrain$Qm * 1000
# Sum up data to get monthly values
library(lubridate)
library(tidyverse)
dayrain$Date=as.POSIXct(dayrain$Date,format="%Y-%m-%d")
# Summarize the dataset by month and year
monthrain=dayrain %>% mutate(Month=month(Date,label = T),Year=year(Date)) %>%
  group_by(Year,Month) %>% summarise(QL=sum(QL,na.rm=T))
```

Now we add in the monthly measurements of Phosphorus concentrations to calculate P loading. We'll start with pool B.
```{r}
# Add in the P dataset (units here are mg/L for TP)
setwd("C:/Users/kande120/OneDrive - Kent State University/Phosphorus Budgets/SJRE/Data")
Pdata<-read.csv("SJRE P Budget Full inletsandoutlets.csv")

# subset out event sampling 
Pdata<- subset(Pdata, Sampling == "baseflow")
# These are the first 6 rows to give you an idea what the datatable looks like
head(Pdata)

# merge the datasets
total <- merge(monthrain,Pdata,by=c("Month","Year"))

# Measure the mass of P entering pool B (units = mg P/year)
total$PinPoolB=total$X5_inlet_poolB*total$QL
# calculate the total mass of P exiting the wetland (units = mg P/year)
total$PoutPoolB=total$X5_Outlet_poolB*total$QL
# Convert to lbs P/year
total$PinlbPoolB=total$PinPoolB/453592
total$PoutlbPoolB=total$PoutPoolB/453592
# calculate the amount of P removed
total$PremovedPoolB=total$PinlbPoolB-total$PoutlbPoolB
RemovalPoolB = sum (total$PremovedPoolB,na.rm=T)
# This is the removal of P by pool by over the course of the 3 months included in this data
print(RemovalPoolB)
# We can make a rough annual estimate by just multiplying by 4:
AnnRemovalPoolB=RemovalPoolB*4
print(AnnRemovalPoolB)
# Calculate percent removal
total$percentPPoolB=total$PremovedPoolB/total$PinlbPoolB
percentRemovalPoolB = mean (total$percentPPoolB,na.rm=T)
print(percentRemovalPoolB)
```


Now lets try the same thing with the entire wetland:
```{r}
# We already have the merged datasets so we start by averaging the various inflows 
total$inflowTP <- rowMeans(total[c('X1_Inlet_N.TP', 'X2_Inlet_NW.TP','X3_Inlet_NE.TP','X18_Inlet_ditchExt.TP.OR.upstream17')], na.rm=TRUE)
# Now calculate the average outflow TP
total$outflowTP <- rowMeans(total[c('X15_Outlet_swale', 'X17_Outlet_ditch.TP')], na.rm=TRUE)
# Measure the mass of P entering the wetland (units = mg P/year)
total$Pin=total$inflowTP*total$QL
# calculate the total mass of P exiting the wetland (units = mg P/year)
total$Pout=total$outflowTP*total$QL
# Convert to lbs P/year
total$Pinlb=total$Pin/453592
total$Poutlb=total$Pout/453592
# calculate the amount of P removed
total$Premoved=total$Pinlb-total$Poutlb
Removal = sum (total$Premoved,na.rm=T)
# This is the removal of P by pool by over the course of the 3 months included in this data
print(Removal)
# We can make a rough annual estimate by just multiplying by 4:
AnnRemoval=Removal*4
print(AnnRemoval)
# Calculate percent removal
total$percentP=total$Premoved/total$Pinlb
percentRemoval = mean (total$percentP,na.rm=T)
print(percentRemoval)
```
### ***ASSUMPTIONS IN BUDGET:*** 
  1. Drainage area is accurate <br>
  2. P isn't lost through groundwater<br>
  3. Baseflow P reduction = storm flow P reduction <br>
  4. Summer P measurements represent annual averages (they don't) <br>
  5. The average of inflow concentrations represents average P concentration entering the wetland   (This likely is NOT true because the amounts of water flowing through each input is different)<br>
  6. Pool B represents the whole wetland <br>

As one last thing lets make a pretty conservative estimate of the error of our budget as the difference between our highest monthly retention and our lowest as an annual average:
```{r}
# First we can take a look at our data: 
total[c("Month","QL","Premoved","percentP")]
```

As you can see there's a big variation in discharge that is skewing our results. Mostly a single month accounts for the whole estimate of P retention.

```{r}
# High retention value (July; we multiply by 9 and add to the existing estimate to create an annual estimate):
Removal_high = ((total[total$Month == "Jul", "Premoved"])*9)+Removal
# Low retention value (June; we multiply by 9 and add to the existing estimate to create an annual estimate):
Removal_low = ((total[total$Month == "Jun", "Premoved"])*9)+Removal
print(Removal_high)
print(Removal_low)
```

This estimate range is taking our 3 measured months as accurate and then assuming that the months we don't know could be either as high as July, or as low as June. 


### Lets see how rainfall data from the closest publicly available weather station does (Ft. Wayne Indiana, 31 miles away)
Calculate drainage area and treated drainage area as above:

```{r}
# Drainage Area calculated by Bishwodeep using Digital Elevation Model (DEM)
tDA = 103.32
# convert tDA to m^2
tDAm2=tDA*4046.86
# units = m^2
print(tDAm2)
```
Then we import data on precipitation from weather underground at the Ft. Wayne weather station (51 miles away)

```{r}
setwd("C:/Users/kande120/OneDrive - Kent State University/Phosphorus Budgets/SJRE/Data")
raindata<-read.csv("Ftwaynerainfall.csv")
# make it recognize that it's a date-time value
raindata = mutate(raindata, datetime = as.POSIXct(datetime, format = "%m/%d/%Y"))
# Rename so our old code keeps working 
dayrain=raindata
```
Then we calculate runoff as above.
```{r}
# Start out by calculating S (inches)
CN=80
S = (1000/CN)-10
# Calculate I (inches)
Ia = 0.2*S
# don't need to convert rain to inches/day, but lets name it so our old code still works
dayrain$precip = dayrain$precipaccum_in
# caculate runoff (inches) for each day in the dataset
dayrain$Q = ifelse(dayrain$precip<=Ia, 0, ((dayrain$precip-Ia)^2)/(dayrain$precip-Ia+S))
# scale up to the entire drainage area
dayrain$Q=tDAm2*dayrain$Q
# convert back to meters 
dayrain$Qm = dayrain$Q * 0.0254
# convert discharge to liters
dayrain$QL = dayrain$Qm * 1000
# Sum up data to get monthly values
library(lubridate)
library(tidyverse)
dayrain$Date=as.POSIXct(dayrain$datetime,format="%Y-%m-%d")
# Summarize the dataset by month and year
monthrain=dayrain %>% mutate(Month=month(Date,label = T),Year=year(Date)) %>%
  group_by(Year,Month) %>% summarise(QL=sum(QL,na.rm=T))
```

Now we add in the monthly measurements of Phosphorus concentrations to calculate P loading.
```{r}
# Add in the P dataset (units here are mg/L for TP)
setwd("C:/Users/kande120/OneDrive - Kent State University/Phosphorus Budgets/SJRE/Data")
Pdata<-read.csv("SJRE P Budget Full inletsandoutlets.csv")
# subset out event sampling 
Pdata<- subset(Pdata, Sampling == "baseflow")
# merge the datasets
total_wayne <- merge(monthrain,Pdata,by=c("Month","Year"))
#  averaging the various inflows 
total_wayne$inflowTP <- rowMeans(total_wayne[c('X1_Inlet_N.TP', 'X2_Inlet_NW.TP','X3_Inlet_NE.TP','X18_Inlet_ditchExt.TP.OR.upstream17')], na.rm=TRUE)
# Now calculate the average outflow TP
total_wayne$outflowTP <- rowMeans(total_wayne[c('X15_Outlet_swale', 'X17_Outlet_ditch.TP')], na.rm=TRUE)
# Measure the mass of P entering the wetland (units = mg P/year)
total_wayne$Pin=total_wayne$inflowTP*total_wayne$QL
# calculate the total_wayne mass of P exiting the wetland (units = mg P/year)
total_wayne$Pout=total_wayne$outflowTP*total_wayne$QL
# Convert to lbs P/year
total_wayne$Pinlb=total_wayne$Pin/453592
total_wayne$Poutlb=total_wayne$Pout/453592
# calculate the amount of P removed
total_wayne$Premoved=total_wayne$Pinlb-total_wayne$Poutlb
Removal = sum (total_wayne$Premoved,na.rm=T)
# This is the removal of P by pool by over the course of the 3 months included in this data
print(Removal)
# We can make a rough annual estimate by just multiplying by 4:
AnnRemoval=Removal*4
print(AnnRemoval)
# Calculate percent removal
total_wayne$percentP=total_wayne$Premoved/total_wayne$Pinlb
percentRemoval = mean (total_wayne$percentP,na.rm=T)
print(percentRemoval)
```

Now lets compare our two datasets:
```{r}
# First our data from out sensor network: 
total[c("Month","QL","Premoved","percentP")]
# Then the dataset using publicly available rainfall data from Ft. Wayne:
total_wayne[c("Month","QL","Premoved","percentP")]
```

Then get the range in our new estimate:

```{r}
# High retention value (July; we multiply by 9 and add to the existing estimate to create an annual estimate):
Removal_high = ((total_wayne[total_wayne$Month == "Jul", "Premoved"])*9)+Removal
# Low retention value (June; we multiply by 9 and add to the existing estimate to create an annual estimate):
Removal_low = ((total_wayne[total_wayne$Month == "Jun", "Premoved"])*9)+Removal
print(Removal_high)
print(Removal_low)
```
